Lonely and online: an investigation of social media use and social isolation in China

### 2024-03-19 tandf/Social Isolation in China.qmd
More updates

While most of the studies agree that the causal direction flows from social media use to feelings of isolation (though see [@nowland2018; @kim2017]), there is disagreement among authors about the specific mechanisms and causal pathways through which this relationship exists. For those who find a negative relationship between social media and well-being, one set of mechanisms suggests that the social media itself that is the problem. Cyberbullying, or more generally, repeated and aggressive online attacks against users has been systematically linked to feelings of social isolation [@ademiluyi2022]. Another proposed mechanism is upward comparisons of others and the weak nature of online connections - both of these have been found to generate social anxiety and social isolation [@alabri2022; @büttner2022]. A third set of mechanisms primarily concern themselves with the physical changes that social media causes. Social media use has been shown to displace face to face activity, which leads to greater social isolation [@larson2018]. Heavy social media use also decreases hours of sleep, leading to greater feelings of anxiety and more social isolation [@twenge2017]. The current research suggests that which of these mechanisms is dominant depends on the age/demographic group in question and the societal context [@twenge2019].

Most of the existing research on the subject focuses on Western highly developed societies, and this focus may limit variation on key independent variables given the similar social and institutional context in which users are embedded. In particular, we believe the case of China can provide leverage in three areas that existing theories have difficulty resolving. The first is that

To address this problem, this study aims to understand how these complex dynamics interact in the non-Western setting of China. Relatively little is known about China's level of social isolation, though what evidence that does exist suggests that it is on the rise. Twenge et al.'s study found that adolescent feelings of loneliness have generally been on the rise in comparable East Asian countries [@twenge2021]. News reports have highlighted the increasing number of people who consider themselves to be online outcasts or losers (*diaosi*) [@nan2013; @szablewicz2014; @chinas2016]. We do not aim to make a direct comparison between China and Western countries but rather highlight the way in which China's unique context helps shed light on which contextual factors matter in understanding the linkage between the two variables.

The most important of these contextual factor in China is that social media space that is highly controlled online environment. Many people are aware that the Chinese government heavily censors politically sensitive content on online platforms. Fewer are aware that the Chinese government spends as much, if not more, effort to clean up non-political negative or distressing content on the Chinese internet, such as scams, instances of cyber bullying, disturbing videos, and vulgar content [@wang2020]. This control impulse also extends content the government views as spiritually polluting, such as pornography, scantily dressed women, and other escapist content. While the Chinese government and online platforms are not 100 per cent effective in their ability to completely stamp out such content, the default is that such content is actively restricted, while US platforms tent to take a more pro-speech regulatory position [@liu2022]. If one of the main mechanisms through which social media promotes social isolation occurs is exposure to negative content, then this highly regulated online environment should lead to a weaker relationship between social media use and isolation. Suggestive of this is a study by Liu and Liu that finds Chinese who consumed more commercialized and sensationalized media reported more trauma during the first COVID-19 outbreak in China than those who followed government sources [@liu2020].

On the other hand, China also has a highly robust online influencer culture. While social media influencers are a global phenomena, China has perhaps the most commercialized version of influencing. Social media influencers are developed by large agencies and work in an almost assembly-line fashion to promote products on livestreams or pitch sponsored products. In this influencer culture, less attention is paid to authenticity and personality and relatively more emphasis is placed on consumerism, brand promotion, and image [@tam2019; @wei2023]. Given that previous research has suggested one of the main channels through which social media creates feelings of isolation is negative personal comparisons with online content (so-called "FOMO" - fear of missing out), heavy social media use may increase the strength of the relationship between social media use and social isolation through consumption of such highly commercialized content. These influencers are most popular among young consumers so one may expect that only younger users would experience these effects.

Additionally, China's social media environment is relatively unique in that the dominant messaging platform is also the largest social media app (WeChat). Users can use it simply to chat with family and friends, they can use it to post photos and other content about their life in a feed that works similarly to Facebook, or they can browse videos of others in a method similar to Tiktok. Previous research has indicated that while, on the whole, social media increases social isolation, for users that primarily use it to contact friends and family, its use can counteract the negative effects [@hajek2019; @lim2020]. In this way, those that heavily use these type of combined platforms may engage in multiple uses of social media that tend to cancel each other out. Given the ubiquity of WeChat in China, such a relationship seems plausible.

Finally, the nexus of social isolation and social media use in China has played out in the context of one of the most rapid periods of urbanization in human history. Hundreds of millions of people have shifted from rural subsistence farming to participating in an urban market economy. The existing literature finds that urbanicity is an important factor in levels and mechanisms for social isolation; many studies of Western societies have found that rural residents, while having stronger familial bonds, also can more easily experience social isolation without those bonds [@kaye2017; @koning2017; @henning-smith2018]. In China, less is known about the differential effects of the urban rural divide on feelings of social isolation but the research that does exist suggests that urban environments can be especially isolating, particularly for rural to urban migrants particularly due to lack of access to traditional support systems and familial connections [@li2017].

Given the existing literature and the context of China, this leads to a series of testable hypotheses.

>>>>

While most of the studies agree that the causal direction flows from social media use to feelings of isolation (though see [@nowland2018; @kim2017]), there is disagreement among authors about the specific mechanisms and causal pathways through which this relationship exists. For those who find a negative relationship between social media and well-being, one set of mechanisms suggests that the social media itself that is the problem. Cyberbullying, or more generally, repeated and aggressive online attacks against users has been systematically linked to feelings of social isolation [@ademiluyi2022]. Another proposed mechanism is upward comparisons of others - users observe content of friends or influencers who post luxurious or exotic experiences which cause users to encounter feelings of unhappiness or FOMO about their own life situation [@alabri2022; @büttner2022]. A third set of mechanisms primarily concern themselves with the physical changes that social media causes. Social media use has been shown to displace face to face activity, which leads to greater social isolation [@larson2018]. Heavy social media use also decreases hours of sleep, leading to greater feelings of anxiety and more social isolation [@twenge2017]. The current research suggests that which of these mechanisms is dominant depends on the age/demographic group in question and the societal context [@twenge2019].

Most of the existing research on the subject focuses on Western highly developed societies, and this focus limits variation on key independent variables given the similar social and institutional context in which users are embedded. To address this problem, this study aims to understand how these complex dynamics interact in the non-Western setting of China. We believe the setting of China gives additional leverage to investigate issues of heterogeneity and the importance of context for several reasons. Firstly, the highly controlled nature of China's internet space prevents user exposure to some of the more threatening online content (such as cyberbullying) that is hypothesized to be a factor in creating a negative relationship between social media and feelings of social isolation [@wang2020]. If this mechanism were the dominant reason for previous findings of a negative relationship between the two, one would expect the relationship to be more positive in China. Relatedly, China has a more commercial and robust influencer culture that often dominates user's social media consumption [@tam2019; @wei2023]. If the main pathway creating a negative relationship is via user's comparisons of their lives with others (FOMO), we expect the relationship between social media use and feelings of social isolation to be greater in China.

Finally, China has undergone a period of very rapid urbanization in the last 40 years, a development that has brought about enormous changes to traditional social structures [@xu2014]. This rapid urbanization can help test two important unresolved questions in the existing literature that are difficult to resolve given that urbanization in Western countries is a process that largely finished by the advent of social media. The first is whether social media can provide a means to decrease social isolation among those who actively seek it out to make friends. China's rapid urbanization and attendant human migration has created a large population of people that lack strong social bonds and may use social media as a method to create new social bonds (as the earlier theorists of social media hypothesized users may employ the technology). The existing literature on the West finds that urbanicity is an important factor in levels and mechanisms for social isolation; many studies of Western societies have found that rural residents, while having stronger familial bonds, also can more easily experience social isolation without those bonds [@kaye2017; @koning2017; @henning-smith2018]. Social media use in China could therefore either help mitigate the feelings of isolation brought on by rapid urbanization by connecting family members who have left the countryside for urban work (based on the earlier, more positive views of social media's potential) or amplify further the feelings of isolation due to several of the proposed negative relationship mechanisms.

The case of China cannot resolve every debate in the literature or test all possible mechanisms but we do believe it can provide insight into a few of the more prominent proposed processes. Thus, given the existing literature and the context of China, we develop a series of testable hypotheses.
